### 2020-12-20 response/response-demography-4.tex

\documentclass[letterpaper,12pt]{article}

% Xetex preamble
\usepackage{fontspec}
\setromanfont[Ligatures=TeX]{TeX Gyre Pagella}
\usepackage{unicode-math}
\setmathfont{TeX Gyre Pagella Math}

\usepackage[longnamesfirst]{natbib}
%\usepackage[tabhead,nolists,tablesfirst]{endfloat}
\usepackage[flushleft]{threeparttable}
\usepackage{booktabs}
\usepackage{rotating}  
\usepackage{dcolumn} 
\usepackage{setspace}
%\usepackage{flafter} 
%\usepackage{longtable}
%\usepackage[pdftex]{graphicx}
\usepackage[xetex,colorlinks=true,linkcolor=black,citecolor=black]{hyperref} 
\usepackage[margin=1.0in]{geometry} 
\usepackage{multirow}

%opening
\title{} \author{}

\doublespacing

\begin{document}

\begin{center} \textbf{\large Birth Spacing in the Presence of Son
Preference and 
Sex-Selective Abortions: India's Experience over Four Decades}
\end{center}

\begin{center} Response to Editor Comments \end{center}

\noindent I have attached the revised version of my paper,
``Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective
Abortions: India's Experience Over Four Decades.''
Thank you for the comments and suggestions. 
Below I outline my response to each issue raised.

The author has undertaken much reorganization and has made very useful clarifications to
benefit the reader in the text and tables. Nevertheless, we still find three issues with
the writing style:

(1) Explanations of key concepts or theories need to be spelled out more clearly,
particularly for the non-economist. Much has been done in this regard, but further work is
still needed.

(2) The arguments do not flow from section to section and paragraph to paragraph. Each
section and argument need to be introduced so the reader understands why this information
is being presented in this place. Many times, the author jumps into new arguments without
an introduction or transition or explanation.

(3) There is no clear outline of the arguments presented in the Background section. Each
argument is presented separately rather than tying into the overall framework and
subsequent analyses.

The following are some examples of (1) and (2); however, the author should check the entire
paper for similar issues and address (3) as well.

1. In general, each argument in the background section is not clear and fully described for
the reader.

a. P. 4 first sentence: The author states that female education is a crucial explanatory
variable and it matters for fertility, sex ratios, mortality, etc. in a list. It would be
useful for the reader to briefly spell out how female education fits into the entire
framework looking across all these outcomes. In addition, the second paragraph on
measurement of education seems like it belongs in the data section, unless there is some
conceptual reason why this information is included here.

b. The 2 paragraphs on female labor force participation on pp. 5-6: The two findings that
support the argument that changes in birth spacing are not due to economic reasons are
merely listed, leaving the reader to make the connection as to why these facts support the
argument herself. The income effect dominating the substitution effect and greater negative
elasticity of women’s wages need to be explained to readers unfamiliar with these concepts
or how they fit together.

c. P. 6 regarding sex selection: “The introduction of sex selection allows parents to avoid
giving birth to girls but increases the expected interval to the next birth.” The reader
needs an explanation here of why sex selection increases birth interval length. The author
mentions waiting time to conception and other components in the introduction (and
footnote). A clear explanation of this argument needs to be reiterated (or moved) here.

d. P. 6: “As women’s education increases, their productivity in the production of offspring
human capital also increases. With relatively more boys born because of increased access to
sexselective abortions and the increasing income potential for (male) offspring, demand for
bettereducated women can increase, even if they do not participate in the labor market.” I
am unclear as to why this argument is included. It needs explanation. Is this related to
length of birth intervals here? Or is this mentioned because women’s education and
increasing demand for better-education women will actually lower the sex ratio?

e. Pp. 6-7: I do not understand the importance of this argument and how it fits into the
authors’ story: “If more and “better” parental attention per child results in higher child
“quality,” we should expect longer birth intervals. However, the evidence on spacing’s
effect on child quality measures such as IQ and education is mixed for developed countries
and nonexisting for developing countries. The exception is health and mortality, where
longer spacing does lead to better outcomes, although this relationship weakens with
maternal education.”

f. P. 7: The final summary paragraph lists predictions based on the earlier arguments. Yet
it is hard to connect these predictions to the paragraphs above. Perhaps predictions could
be noted with the earlier arguments.

2. Estimation strategy section.

In the first paragraph, it would be useful to clearly outline the analyses that will be
undertaken (such as first, I document changes in birth intervals over time and how
influenced by sex selection, second how birth intervals affected fertility, and third how
birth intervals affected mortality). As it stands, this section jumps into a discussion of
the hazard model and the reader does not know in which analyses it will be used.

Some additional issue as examples where more clarification is needed:

3. Abstract. Please reorganize to start with a statement of the problem and main aims of
the study. As it stands, the abstract is a list of findings, with no context as to the
issues/questions the author investigates. Please be cognizant of using terms such as “women
most likely to use sexselective abortion.” The reader does not understand how this is
measured yet, and may get the false impression that sex-selection abortion is measured
directly.

4. P. 1: The second motivation is unclear. “the combined changes in birth spacing may
outpace what we have observed in other countries.” Outpaced means what? decreasing faster,
or increasing faster, or sex selection plays more of a role?

5. Throughout the paper, the author compares the total fertility rate to cohort fertility
rate. For the former, is the period fertility rate a more accurate definition?

6. P. 2: the main point of this sentence is unclear: “Counteracting effect is possible if
longer birth intervals arise from multiple abortions because the short duration between
pregnancies could increase mortality.”

7. Sometimes the use of the term “birth interval” actually refers to “birth interval
length.” Please check the uses of these terms and correct accordingly.

8. P. 2: “The key variables are maternal education, the sex of previous children, and the
area of residence.” Does the author mean key independent or predictor variables?

9. The author should clearly note that sex selection is not observed and tone down some
conclusions, such as “Sex selection, however, is behind the most substantial increases in
birth spacing. The best-educated women with two girls had the most biased sex ratio and the
most significant increase in birth intervals.” The author could say that sex selection
appears to be behind these findings; or evidence suggests that sex selection is behind
these findings, etc. In addition, clarify that those most likely to practice sex selection
(women with highest education and two girls) is based on your analysis, not settled in the
literature.

10. While it might be messy to show the confidence intervals in all the figures, is it
possible to note some significant differences in the text?

\newpage
\bibliographystyle{econometrica}
\bibliography{../paper/sex_selection_spacing.bib}
\addcontentsline{toc}{section}{References}

\end{document}

### 2020-12-21 paper/sex_selection_spacing-ver5.tex
Begin final round of editing

\noindent India has experienced an increasingly male-biased sex ratio at birth following
the introduction of prenatal sex determination techniques. 
I examine three questions.
First, how did birth spacing respond to this and other secular developments? 
Second, did changes in birth spacing make us overestimate the decline in cohort fertility? 
Finally, did the advent of sex-selective abortions change the relationship between birth 
spacing and infant mortality? 
Well-educated Hindu women with no sons had the most substantial lengthening of birth 
intervals and the most male-biased sex ratios, likely from sex-selective abortions. 
Consequently, some women with no sons now have longer birth intervals than those with sons, 
reversing India's traditional spacing pattern.
Less-educated women continued short spacing when they have girls, with only limited
evidence of male-biased sex ratios. 
Because of the overall lengthening in spacing, period fertility rates substantially 
overestimated how fast cohort fertility fell. 
Furthermore, predicted cohort fertility remains above replacement for all but the 
best-educated urban women. 
If lengthening of birth intervals occurs from repeated abortions, the associated
short pregnancy spacing may counteract any positive effects of longer birth spacing. 
There is no evidence of this effect on infant mortality. 
Finally, judging from sex ratios, sex-selective abortion use is not declining.

@@

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, past research has failed to appreciate that the growing use of sex-selective 
abortions may substantially increase birth spacing; as detailed below, it takes 6 months 
or more after an abortion to reach the same point in the next pregnancy.
Second, women's greater educational attainment, higher household income, but low and 
declining female labor force participation all likely influence birth spacing. 
The combination of the increasing use of sex selection and the other societal changes 
can significantly impact birth spacing, but we know little about how much.

@@

First, did changes in birth spacing make us overestimate the decline in cohort fertility?
With longer spacing, mothers will be older at each parity, and this tempo effect 
makes period fertility measures, such as the total fertility rate, downward-biased
estimates of cohort fertility \citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
Hence, a rapidly expanding use of sex selection could make the total fertility rate---our 
most used fertility measure---fall substantially faster than cohort fertility.
In this case, households' fertility goals may be higher than generally accepted.

Second, what is the relationship between infant mortality and the changes in birth spacing
and sex selection? 
In India, birth intervals have traditionally been shorter with fewer sons, contributing to 
girls' higher mortality risk
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}. 
Therefore, longer birth spacing may reduce mortality through, for example, diminished 
sibling competition \citep{Conde-Agudelo2012,Molitoris2019}. 
However, if the spacing between births lengthens because of sex-selective abortions, the 
spacing between pregnancies may still be very short. 
Short pregnancy spacing may lead to worse child outcomes because of maternal nutritional 
depletion and insufficient time to recover from the previous pregnancy. 
Hence, children born after long birth intervals where multiple abortions have punctuated 
those intervals may not see the same benefits as a child born after a long interval that 
is not punctuated by abortions.

@@

The primary outcomes I examine are the 25th, 50th, and 75th percentile birth intervals lengths;
the sex ratio at birth; and the likelihood of giving birth. 
I estimate the model across four periods to capture the changing access and 
legality of sex selection. 
The key independent variables are maternal education, the sex of previous children, and 
the area of residence.

@@

First, birth intervals lengthened over the four decades, with the lengthening longer, the
higher the parity, the more educated the woman, and the higher the percentile.

First, birth intervals lengthened over the four decades, with the lengthening longer, the 
higher the parity, the more educated the woman, and the higher the percentile. 
Well-educated women with no sons had the most substantial lengthening of birth intervals 
and the most male-biased sex ratios, both likely arising from sex-selective abortions. 
Consequently, some women with no sons now have \emph{longer} birth intervals than those 
with sons, reversing India's traditional spacing pattern. 
Less-educated women continued with short spacing when they have girls, with only limited 
evidence of male-biased sex ratios.
The likelihood of a very short birth interval changed little across most groups.

Second, the period fertility rate substantially overestimated how fast cohort fertility 
fell in the 1990s and early 2000s as spacing began to increase. 
Although the two have lately been converging, the predicted cohort fertility is still 
10\%--20\% higher than the period fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

@@

This section provides a conceptual framework for understanding changes in birth spacing.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I discuss why female education and area of residence are the principal factors 
in the empirical analyses and how they tie into the three explanations.

@@

% Child outcomes
With increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means a stronger incentive to invest in their children's
education, which increases the cost of having children, thereby lowering fertility. 
With lower expected mortality, parents need fewer births to reach the desired number of
surviving children, allowing parents to invest more in each child.

An increased desire for better child outcomes will also tend to lengthen birth spacing,
provided it has positive effects on child outcomes.
The clearest example of a positive effect of longer spacing that very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

@@

>>>>

\noindent India has experienced an increasingly male-biased sex ratio at birth following
the introduction of prenatal sex determination techniques. 
I examine three questions.
First, how did birth spacing change with this and other secular developments? 
Second, did changes in birth spacing make us overestimate the decline in cohort fertility? 
Finally, did the advent of sex-selective abortions change the relationship between birth 
spacing and infant mortality? 
Well-educated Hindu women with no sons had the most substantial lengthening of birth 
intervals and the most male-biased sex ratios, likely from sex-selective abortions. 
Consequently, some women with no sons now have longer birth intervals than those with sons, 
reversing India's traditional spacing pattern.
Less-educated women continued short spacing when they have girls, with only limited
evidence of male-biased sex ratios. 
Because of the overall lengthening in spacing, period fertility rates substantially 
overestimated how fast cohort fertility fell. 
Furthermore, predicted cohort fertility remains above replacement for all but the 
best-educated urban women. 
If lengthening of birth intervals occurs from repeated abortions, the associated
short pregnancy spacing may counteract any positive effects of longer birth spacing. 
There is no evidence of this effect on infant mortality. 
Finally, judging from sex ratios, sex-selective abortion use is not declining.

@@

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, past research has failed to appreciate that the use of sex-selective abortions could 
substantially increase birth spacing. 
As detailed below, it takes 6 months or more after an abortion to reach the same point in 
the next pregnancy. 
Second, other societal changes, such as women's greater educational attainment, higher 
household income, and a low and declining female labor force participation, all 
likely influence birth spacing. 
The combination of the apparent increasing use of sex selection and the other societal 
changes can significantly impact birth spacing, but we know little about how much.

@@

First, did changes in birth spacing lead us to overestimate the decline in cohort fertility?
With longer spacing, mothers will be older at each parity, and this tempo effect 
makes period fertility measures, such as the total fertility rate, downward-biased
estimates of cohort fertility \citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
Hence, a rapidly expanding use of sex selection could make the total fertility rate---our 
most used fertility measure---fall substantially faster than cohort fertility.
In this case, households' fertility may be higher than generally accepted.

Second, what is the relationship between infant mortality and the changes in birth spacing
and sex selection? 
In India, birth intervals have traditionally been shorter with fewer sons, contributing to 
girls' higher mortality risk
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}. 
Therefore, longer birth spacing may reduce mortality through, for example, diminished 
sibling competition \citep{Conde-Agudelo2012,Molitoris2019}. 
However, if the spacing between births lengthens because of sex-selective abortions, the 
spacing between pregnancies may still be very short. 
Short pregnancy spacing may lead to worse child outcomes because of maternal nutritional 
depletion and insufficient time to recover from the previous pregnancy. 
Hence, children born after long birth intervals where multiple abortions have punctuated 
those intervals may not see the same benefits as children born after a long interval that 
is not punctuated by abortions.

@@

The primary outcomes I examine are the 25th, 50th, and 75th percentile birth intervals 
durations;
the sex ratio at birth; and the likelihood of giving birth. 
I estimate the model across four periods to capture the changing access and 
legality of sex selection. 
The key explanatory variables are maternal education, the sex of previous children, and 
the area of residence.

@@

First, birth intervals lengthened over the four decades, with the lengthening longer, the
higher the parity, the more educated the woman, and the higher the percentile.
Well-educated women with no sons had the most substantial lengthening of birth intervals 
and the most male-biased sex ratios, both likely arising from sex-selective abortions. 
Consequently, some women with no sons now have \emph{longer} birth intervals than those 
with sons, reversing India's traditional spacing pattern. 
Less-educated women continued to have short spacing when they have girls, with only limited 
evidence of male-biased sex ratios.
The likelihood of a very short birth interval changed little across most groups.

Second, the period fertility rate substantially overestimated how fast cohort fertility 
fell in the 1990s and early 2000s as spacing initially increased. 
Although the two have lately been converging, the predicted cohort fertility is still 
10\%--20\% higher than the period fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

@@

This section provides a conceptual framework for understanding changes in birth spacing.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, investment in children, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I discuss why female education, area of residence, and sex composition are the
principal factors in the empirical analyses and tie them into the three explanations.

@@

% Investment in children
With increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means a stronger incentive to invest in their children's
education, thereby increasing the cost of having children and lowering fertility. 
With lower expected mortality, parents need fewer births to reach the desired number of
surviving children, which, in turn, allows parents to invest more in each child.

An increased desire to invest in children may also lengthen birth spacing, provided it has 
positive effects on child outcomes. 
The clearest example of a positive effect of longer spacing is for health as very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

@@

[HERE!]
